---
title: "Practical Machine Learning Writeup"
author: "chhavi"
date: "Friday, August 14, 2015"
output: html_document
---
We first download the data from the links provided and copy them to our working directory , interpreting the miscellaneous NA, #DIV/0! and empty fields as NA:

```{r}
training <- read.csv("pml-training.csv", na.strings = c("NA", "#DIV/0!", ""))
testing  <- read.csv("pml-testing.csv",  na.strings = c("NA", "#DIV/0!", ""))
dim(training)
str(training, list.len=10)
table(training$classe)
table(training$user_name)
prop.table(table(training$user_name, training$classe), 1)
prop.table(table(training$classe))

```
Based on the above information, let's first do some basic data clean-up by removing columns 1 to 6, which are there just for information and reference purposes:

```{r}
training <- training[, 7:160]
testing  <- testing[, 7:160]
```

and removing all columns that are mostly NA:

```{r}
is_data  <- apply(!is.na(training), 2, sum) > 19621
training <- training[, is_data]
testing  <- testing[, is_data]


```
Before we can move forward with data analysis, we split the training set into two for cross validation purposes. We randomly subsample 60% of the set for training purposes (actual model building), while the 40% remainder will be used only for testing, evaluation and accuracy measurement.

Loadign required package

```{r}
library(caret)
library(ggplot2)

```

```{r}
set.seed(3141592)
inTrain <- createDataPartition(y=training$classe, p=0.60, list=FALSE)
train1  <- training[inTrain,]
train2  <- training[-inTrain,]
dim(train1)
dim(train2)
```

We can now [i] identify the "zero covariates"" from train1 and [ii] remove these "zero covariates"" from both train1 and train2:

```{r}
nzv_cols <- nearZeroVar(train1)
if(length(nzv_cols) > 0) {
  train1 <- train1[, -nzv_cols]
  train2 <- train2[, -nzv_cols]
}
dim(train1)
dim(train2)

```
##Data Manipulation
53 covariates is a lot of variables. Let's look at their relative importance using the output of a quick Random Forest algorithm (which we call directly using randomForest() rather than the caret package purely for speed purposes as we cannot specify the number of trees to use in caret), and plotting data importance using varImpPlot():

```{r}
library(randomForest)
set.seed(3141592)
fitModel <- randomForest(classe~., data=train1, importance=TRUE, ntree=100)
varImpPlot(fitModel)
```

Using the Accuracy and Gini graphs above, we select the top 10 variables that we'll use for model building. If the accuracy of the resulting model is acceptable, limiting the number of variables is a good idea to ensure readability and interpretability of the model. A model with 10 parameters is certainly much more user friendly than a model with 53 parameters.

Our 10 covariates are: yaw_belt, roll_belt, num_window, pitch_belt, magnet_dumbbell_y, magnet_dumbbell_z, pitch_forearm, accel_dumbbell_y, roll_arm, and roll_forearm.
...........................................................................................

Let's analyze the correlations between these 10 variables. The following code calculates the correlation matrix, replaces the 1s in the diagonal with 0s, and outputs which variables have an absolute value correlation above 75%:

```{r}
correl = cor(train1[,c("yaw_belt","roll_belt","num_window","pitch_belt","magnet_dumbbell_z","magnet_dumbbell_y","pitch_forearm","accel_dumbbell_y","roll_arm","roll_forearm")])
diag(correl) <- 0
which(abs(correl)>0.75, arr.ind=TRUE)
```

So we may have a problem with roll_belt and yaw_belt which have a high correlation (above 75%) with each other:

```{r}
cor(train1$roll_belt, train1$yaw_belt)
```

after plotting so many plots between 2 diffrent variable  identified an interesting relationship between roll_belt and magnet_dumbbell_y:


```{r}
qplot(roll_belt, magnet_dumbbell_y, colour=classe, data=train1)
```

This graph suggests that one could probably categorize the data into groups based on roll_belt values.

..................................................................................

##Modeling

We are now ready to create our model.
We are using a Random Forest algorithm, using the train() function from the caret package.
We are using 9 variables out of the 53 as model parameters. These variables were among the most significant variables generated by an initial Random Forest algorithm, and are roll_belt, num_window, pitch_belt, magnet_dumbbell_y, magnet_dumbbell_z, pitch_forearm, accel_dumbbell_y, roll_arm, and roll_forearm. These variable are relatively independent as the maximum correlation among them is 50.57%.
We are using a 2-fold cross-validation control. This is the simplest k-fold cross-validation possible and it will give a reduced computation time. Because the data set is large, using a small number of folds is justified.

```{r,   Cache = TRUE,eval= FALSE}
set.seed(3141592)
fitModel <- train(classe~roll_belt+num_window+pitch_belt+magnet_dumbbell_y+magnet_dumbbell_z+pitch_forearm+accel_dumbbell_y+roll_arm+roll_forearm,
                  data=train1,
                  method="rf",
                  trControl=trainControl(method="cv",number=2),
                  prox=TRUE,
                  verbose=TRUE,
                  allowParallel=TRUE)
```

so we may want to save the model generated for later use

```{r,eval=FALSE}
saveRDS(fitModel, "modelRF.Rds")
```

We can later use this tree, by allocating it directly to a variable using the command:

```{r}
fitModel <- readRDS("modelRF.Rds")
```


###How accurate is this model?
We can use caret's confusionMatrix() function applied on train2 (the test set) to get an idea of the accuracy:

```{r}
predictions <- predict(fitModel, newdata=train2)
confusionMat <- confusionMatrix(predictions, train2$classe)
confusionMat
```
99.77% is a very impressive number for accuracy which totally validates the idea / hypothesis we made to eliminate most variables and use only 9 relatively independent covariates.

The out-of-sample error rate is 0.23%.

##Estimation of the out-of-sample error rate

The train2 test set was removed and left untouched during variable selection, training and optimizing of the Random Forest algorithm. Therefore this testing subset gives an unbiased estimate of the Random Forest algorithm's prediction accuracy (99.77% as calculated above). The Random Forest's out-of-sample error rate is derived by the formula 100% - Accuracy = 0.23%, or can be calculated directly by the following lines of code:

```{r}
missClass = function(values, predicted) {
  sum(predicted != values) / length(values)
}
OOS_errRate = missClass(train2$classe, predictions)
OOS_errRate
```

##Coursera Submission
We predict the classification of the 20 observations of the testing data set for Coursera's "Course Project: Submission" challenge page:

```{r}
predictions <- predict(fitModel, newdata=testing)
testing$classe <- predictions
```
We create one .CSV file with all the results, presented in two columns (named problem_id and classe) and 20 rows of data:

```{r}
submit <- data.frame(problem_id = testing$problem_id, classe = predictions)
write.csv(submit, file = "coursera-submission.csv", row.names = FALSE)
```

And we create twenty .TXT file that we will upload one by one in the Coursera website (the 20 files created are called problem_1.txt to problem_20.txt):

```{r}
answers = testing$classe
write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_",i,".txt")
    write.table(x[i], file=filename, quote=FALSE, row.names=FALSE, col.names=FALSE)
  }
}
write_files(answers)
```



##Conclusion
In this assignment, we accurately predicted the classification of 20 observations using a Random Forest algorithm trained on a subset of data using less than 20% of the covariates.

The accuracy obtained (accuracy = 99.77%, and out-of-sample error = 0.23%) is obviously highly suspicious as it is never the case that machine learning algorithms are that accurate, and a mere 85% if often a good accuracy result.

Either the 6 participants for whom we have data were extraordinarily obedient (for more than 19 thousand observations, a strong performance! This however might be explained by the highly controlled conditions of the data collection), or the data was somehow doctored for this class, or additional testing needs to be performed on other different participants, or Fitbit really works!

It may be interesting to apply the fitModel tree from the Random Forest algorithm obtained in this paper (without any re-calibration) to a completely new set of participants, to complement and validate the analysis.

This project was a very interesting introduction to practical machine learning, and opened up many doors in machine learning in R.
